---
title: Accessing data from the Human Cell Atlas (HCA)
teaching: 20 # Minutes of teaching in the lesson
exercises: 10 # Minutes of exercises in the lesson
---

:::::::::::::::::::::::::::::::::::::: questions 

- How to obtain single-cell reference maps from the Human Cell Atlas?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Learn about different resources for public single-cell RNA-seq data.
- Access data from the Human Cell Atlas using the `CuratedAtlasQueryR` package.
- Query for cells of interest and download them into a `SingleCellExperiment` object. 

::::::::::::::::::::::::::::::::::::::::::::::::


# Single Cell data sources

## HCA Project

The Human Cell Atlas (HCA) is a large project that aims to learn from and map
every cell type in the human body. The project extracts spatial and molecular
characteristics in order to understand cellular function and networks. It is an
international collaborative that charts healthy cells in the human body at all
ages. There are about 37.2 trillion cells in the human body. To read more about
the project, head over to their website at https://www.humancellatlas.org.

## CELLxGENE

CELLxGENE is a database and a suite of tools that help scientists to find,
download, explore, analyze, annotate, and publish single cell data. It includes
several analytic and visualization tools to help you to discover single cell
data patterns. To see the list of tools, browse to
https://cellxgene.cziscience.com/.

## CELLxGENE | Census

The Census provides efficient computational tooling to access, query, and
analyze all single-cell RNA data from CZ CELLxGENE Discover. Using a new access
paradigm of cell-based slicing and querying, you can interact with the data
through TileDB-SOMA, or get slices in AnnData or Seurat objects, thus
accelerating your research by significantly minimizing data harmonization at
https://chanzuckerberg.github.io/cellxgene-census/.

## cellNexus

[`cellNexus`](https://cellnexus.org/) is "a query interface for programmatic
exploration and retrieval of harmonised, curated, and reannotated CELLxGENE
human-cell-atlas data." This is what we'll be using in this lesson for the most
part since having the data pre-harmonised and pre-annotated makes life simpler.

## Data Sources in R / Bioconductor

There are a few options to access single cell data with R / Bioconductor.

| Package | Target | Description |
|---------|-------------|---------|
| [cellxgenedp](https://bioconductor.org/packages/cellxgenedp) | [CellxGene](https://cellxgene.cziscience.com/) | Human and mouse SC data including HCA |
| [cellNexus](https://cellnexus.org/) | [CellxGene](https://cellxgene.cziscience.com/) | fine-grained query capable CELLxGENE data including HCA |

## Installation

If you don't have `cellNexus` already:

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("cellNexus")
```

## Package load 

```{r chunk-opts, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, message = FALSE, warning = FALSE)
library(BiocStyle)
```

```{r}
library(cellNexus)
library(dplyr)
```

## HCA Metadata

The metadata allows the user to get a lay of the land of what is available
via the package. In this example, we are using the sample database URL which
allows us to get a small and quick subset of the available metadata.

```{r, message = FALSE}
sample_url <- cellNexus::SAMPLE_DATABASE_URL

metadata <- get_metadata(cloud_metadata = sample_url) |> 
  collect()
```

Some database details: `get_metadata()` returns a "connection" to the duckdb
server hosting the sample metadata, so we used the `collect()` function to pull
the corresponding table into our R session as a data.frame. This is fine for the
small sample database, but for larger tables with huge numbers of rows, it's
generally better to run a filtered query on the connection and *then* collect
the much smaller result.

Get a view of the first 10 columns in the metadata with `glimpse()`

```{r}
metadata |>
  select(1:10) |>
  glimpse()
```

These are just the first ten, but there are many more metadata columns as we'll
see.

## A tangent on the pipe operator

The vignette materials provided by `CuratedAtlasQueryR` show the use of the
'native' R pipe (implemented after R version `4.1.0`). For those not familiar
with the pipe operator (`|>`), it allows you to chain functions by passing the
left-hand side as the first argument to the function on the right-hand side. It is used extensively in the `tidyverse` dialect of R, especially within the [`dplyr` package](https://dplyr.tidyverse.org/).

The pipe operator can be read as "and then". Thankfully, R doesn't care about whitespace, so it's common to start a new line after a pipe. Together these points enable users to "chain" complex sequences of commands into readable blocks.

In this example, we start with the built-in `mtcars` dataset and then filter to rows where `cyl` is not equal to 4, and then compute the mean `disp` value by each unique `cyl` value.

```{r}
mtcars |> 
  filter(cyl != 4) |> 
  summarise(avg_disp = mean(disp),
            .by = cyl)
```

This command is equivalent to the following:

```{r eval=FALSE}
summarise(filter(mtcars, cyl != 4), avg_disp = mean(disp), .by = cyl)
```

## Exploring the metadata

Let's examine the metadata to understand what information it contains.

We can tally the tissue types across datasets to see what tissues the experimental data come from:

```{r}
metadata |>
  distinct(tissue_groups, dataset_id) |> 
  count(tissue_groups) |> 
  arrange(-n)
```

```{r include=FALSE}
tissue_cnts <- metadata |>
  distinct(tissue_groups, dataset_id) |> 
  count(tissue_groups) |> 
  arrange(-n)
```


That is to say, there are `r tissue_cnts$n[1]` studies that investigate `r tissue_cnts$tissue_groups[1]`.

We can do the same for the imputed ethnicities:

```{r}
metadata |>
    distinct(imputed_ethnicity, dataset_id) |>
    count(imputed_ethnicity)
```

:::: challenge

Look at the other metadata columns with `colnames(metadata)` and inspect a few
that catch your interest.

::: solution

Let's look at filesize, sex, and cell_type. We'll collect the results locally
and shuffle the rows here just to see some variability beyond the first sample
listed.

```{r}
metadata |> 
  select(age_days, cell_type_unified_ensemble) |> 
  slice_sample(prop = 1)
```

You can see that age_days is commonly NA and cell types are mostly immune
related (that's what was selected for in the sample database).

:::

::::

## Downloading single cell data 

The data can be provided as either "counts" or counts per million "cpm" as given
by the `assays` argument in the `get_single_cell_experiment()` function. By
default, the `SingleCellExperiment` provided will contain only the 'counts'
data.

For the sake of demonstration, we'll focus this small subset of samples. We use the `filter()` function from the `dplyr` package to identify cells meeting the following criteria:

* African ethnicity
* any 10x assay
* lung parenchyma tissue
* CD4 cells

<!-- TODO: Find an example that works better with the sample database  -->

```{r}
sample_subset <- metadata |>
    filter(
        grepl("East Asian", imputed_ethnicity) &
        tissue == "lung parenchyma" 
    )
```

Out of the `r nrow(metadata)` cells in the sample database, `r nrow(sample_subset)` cells meet this criteria.

Now we can use `get_single_cell_experiment()`:

```{r, message = FALSE}
single_cell_counts <- sample_subset |>
    get_single_cell_experiment()

single_cell_counts
```

You can provide different arguments to `get_single_cell_experiment()` to get different formats or subsets of the data, like data scaled to counts per million:

```{r, message = FALSE}
sample_subset |>
  get_single_cell_experiment(assays = "cpm")
```

or data on only specific genes:

```{r, message = FALSE}
single_cell_counts <- sample_subset |>
    get_single_cell_experiment(assays = "cpm", features = "PUM1")

single_cell_counts
```

Or if needed, the H5 `SingleCellExperiment` can be returned a Seurat
object (note that this may take a long time and use a lot of memory depending on
how many cells you are requesting).

```{r,eval=FALSE}
single_cell_counts <- sample_subset |>
    get_seurat()

single_cell_counts
```

## Save your `SingleCellExperiment`

Once you have a dataset you're happy with, you'll probably want to save it. The recommended way of saving these `SingleCellExperiment` objects is to use
`saveHDF5SummarizedExperiment` from the `HDF5Array` package.

```{r, eval=FALSE}
single_cell_counts |> saveHDF5SummarizedExperiment("single_cell_counts")
```

## Exercises

:::::::::::::::::::::::::::::::::: challenge

#### Exercise 1: Basic counting + piping

Use `count` and `arrange` to get the number of cells per tissue in descending
order.

:::::::::::::: solution

```{r}
metadata |>
    count(tissue) |>
    arrange(-n)
```
:::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::: challenge

#### Exercise 2: Tissue & type counting

`count()` can group by multiple factors by simply adding another grouping column
as an additional argument. Get a tally of the highest number of cell types per
tissue combination. What tissue has the most numerous type of cells?

:::::::::::::: solution

```{r}
metadata |>
    count(tissue, cell_type) |>
    arrange(-n) |> 
    head(n = 1)

```
:::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::: challenge

#### Exercise 3: Comparing metadata categories

Spot some differences between the `tissue` and `tissue_harmonised` columns.
Use `count` to summarise.

:::::::::::::: solution

```{r}
metadata |>
    count(tissue) |>
    arrange(-n)

metadata |>
    count(tissue_harmonised) |>
    arrange(-n)
```

For example you can see that `tissue_harmonised` merges the `cortex of kidney`
and `kidney` groups in `tissue`.

To see the full list of curated columns in the metadata, see the Details section
in the `?get_metadata` documentation page.
    
:::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::: challenge

#### Exercise 4: Highly specific cell groups

Now that we are a little familiar with navigating the metadata, let's obtain
a `SingleCellExperiment` of 10X scRNA-seq counts of `cd8 tem` `lung` cells for
females older than `80` with `COVID-19`. Note: Use the harmonized columns, where
possible. 

:::::::::::::: solution

```{r, message = FALSE}
metadata |> 
    filter(
        sex == "female" &
        age_days > 80 * 365 &
        grepl("10x", assay) &
        disease == "COVID-19" &  
        tissue_harmonised == "lung" & 
        cell_type_harmonised == "cd8 tem"
    ) |>
    get_single_cell_experiment()
```

You can see we don't get very many cells given the strict set of conditions we used.
:::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: keypoints 

- The `CuratedAtlasQueryR` package provides programmatic access to single-cell reference maps from the Human Cell Atlas.
- The package provides functionality to query for cells of interest and to download them into a `SingleCellExperiment` object.

::::::::::::::::::::::::::::::::::::::::::::::::


## Session Info

```{r, tidy=TRUE}
sessionInfo()
```
